Hippocampal replay of experience at real world speeds

### 2021-05-13 Article.tex

In the context a rodent navigating through space preferentially fire at specific locations in an environment, and thus ensembles of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and thus "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

@@

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approaches for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data, where the shuffled version of the data represent an implicit definition of a "random" sequence.

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% of all SWRs include spatially coherent structure and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs have time points with spatially incoherent structure (\FIG{3}G). To further validate this result, we performed a shuffle that randomized the order of the reward well to reward well runs and then circularly permuted the resulting data. This, in contrast to our previous shuffle, will preserve the local correlations between spikes but reduce global spatial structure. Using this shuffle, we found that there was still less spatially coherent and more spatially incoherent content compared to the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup3}).

@@

We note here that while the size of the HPD region will be influenced by the dynamic estimated from the data, it remains possible for continuous or stationary dynamics to correspond to a large HPD region (see outliers in \FIG{4}E), indicating less spatial certainty for those specific events. In general, if the data does not exhibit strong place specific firing, the HPD region will be large, regardless of dynamic classification. To show this, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes have much larger HPD regions than the real data (recording session \#1: 55 cm vs. 229 cm, p=0.02, real vs. shuffled; recording session \#2: 72 cm vs. 231 cm, p=0.02, real vs. shuffled; one-sided Mann-Whitney-U, \FIGSUPP[4]{fig4sup1}). We also performed a different shuffle that preserved more of the local spatial structure while destroying global spatial structure by permuting the order of the well-to-well runs and circularly shifting the position within the runs. This also resulted in larger HPD regions than the real data (recording session \#1: p=0.02, recording session \#2: p=0.02, \FIGSUPP[4]{fig4sup1}).

@@

% Figure 6
\begin{figure}
\includegraphics[width=0.80\linewidth]{figures/Figure6/Figure6_v2}
\caption{
Comparison of the state space decoder to the standard "Bayesian" decoder. For each panel, the top five rows are the probability density of the estimated speed for SWRs that contain that dynamic or combination of dynamics. Note that these categories are not mutually exclusive since a SWR can have more than one dynamic or combination of dynamics. The sixth row is all SWRs containing unclassified dynamics. The final row is the estimated speed for all SWRs. \textbf{(A)} Radon transform  \textbf{(B)} MAP estimate  \textbf{(C)} Linear regression  \textbf{(D)} State space decoder presented in this manuscript.
}
\label{fig:6}

\end{figure}
\subsection*{Control analyses}

@@

We then asked whether the slow spatially coherent events (stationary-continuous mixtures) potentially reflect a slow theta sweep rather than a replay event. We noted that this seems very unlikely given that the stationary-continuous mixtures tended to represent locations far from the animal that would only be seen during rapid theta sequences (median 52 cm \FIG{5}C). In addition, the majority of our events occur when the animal is moving at speeds < 2 cm/s, as expected for SWRs (\FIGSUPP[5]{fig5sup3}B). Further, when we restrict the analyses to SWRs where the animal was moving at speeds < 1 cm/s, we find similar proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}C).

@@

\subsection*{Comparison to standard approaches}
Finally, we asked how our method compares to the standard approaches to identifying significant events and characterizing their content. This comparison is complicated due to the variety of parameters and inference methods used previously, so here we chose three commonly used approaches and a reasonable parameter set for the comparison. We began with an examination of events based on a measure of statistical significance, which has been the standard in the field.  We applied the clusterless decoder in 20 ms bins to produce probability densities of represented locations, as this binning is comparable to previous methods. We then used the radon transform method \citep{DavidsonHippocampalReplayExtended2009, FarooqEmergencepreconfiguredplastic2019} to calculate, for each SWR, a measure of significance based on the assumptions that the probability density moves at a constant speed throughout the event. We found that ~12\% of the classified events (using the full state-space model) were identified as significant. Not surprisingly, when events were classified based on which dynamic they contained, we found substantial variability: of 7176 SWRs that contained a period of stationary dynamics, 542 (8\%) were significant while of 4511 SWRs that contained a period of continuous dynamics, 1386 (~30\%) were significant. And when we restricted the analysis to stationary events that were >100 ms long, we found that 28\% were significant, and increase that is not surprising given that the shuffle used for the radon method penalizes shorter events. We also note that the linear regression method \citep{KarlssonAwakereplayremote2009, CarrHippocampalreplayawake2011, ShinDynamicsAwakeHippocampalPrefrontal2019}, detects only non-zero slopes as significant, and thus cannot identify stationary events as significant. The substantial variability in the proportions of "significant" events indicates that conclusions about which events are meaningful to the brain could be very different depending on the specific method and parameters used.

@@

We hypothesize that this large set of events supports the ability to store and update  representations of individual places and snippets of experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{YuDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to store or update the value of specific locations or short movement trajectories without having to generate full trajectories representing one of multiple ways to arrive at a location \citep{YuHippocampalcorticalinteraction2015}. Events that contain multiple dynamics (e.g. two continuous representations separated by a brief fragmented representation) might also help bind together experiences that occur in the same environment but at different places. Alternatively, it might be that downstream structures are attuned to these dynamics, and that each period of coherent hippocamapal dynamics drives its own set of responses elsewhere in the brain. We further note that recent results from a study of activity patterns during movement indicates that individual theta cycles can have different dynamics as well \citep{WangAlternatingsequencesfuture2020}, suggesting that the underlying sequences seen during replay and theta may overlap in their content and dynamics.

>>>>

In the context of rodents navigating through space preferentially fire at specific locations in an environment, and thus ensembles of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and thus "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

@@

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approaches for detecting sequential replay involve fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data, where the shuffled version of the data represent an implicit definition of a "random" sequence.

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods and \FIG{1}F). We find that 96\% (22,170 of 23,071) of classified SWRs and 86\% of all SWRs include spatially coherent structure and that this prevalence of spatially coherent structure is consistent across animals (\FIG{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We find that only 14\% (3,295 of 23,071) of classified SWRs have time points with spatially incoherent structure (\FIG{3}G). To further validate this result, we performed a shuffle that randomized the order of the reward well to reward well runs and then circularly permuted the resulting data. This, in contrast to our previous shuffle, will preserve the local correlations between spikes but reduce global spatial structure. Using this shuffle, we found that there was still less spatially coherent and more spatially incoherent content compared to the real data (recording session \#1, spatially coherent: 99\% vs. 71\%, p=0.02, real vs. shuffled; recording session \#1, spatially incoherent: 22\% vs. 61\%, p=0.02, real vs. shuffled; recording session \#2, spatially coherent: 95\% vs. 63\%, p=0.02, real vs. shuffled; recording session \#2, spatially incoherent: 16\% vs. 79\%, p=0.02, real vs. shuffled; \FIGSUPP[3]{fig3sup3}).

@@

We note here that while the size of the HPD region will be influenced by the dynamic estimated from the data, it remains possible for continuous or stationary dynamics to correspond to a large HPD region (see outliers in \FIG{4}E), indicating less spatial certainty for those specific events. In general, if the data does not exhibit strong place specific firing, the HPD region will be large, regardless of dynamic classification. To show this, we used the same shuffle as above, resampling position with replacement for two recording sessions and shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes have much larger HPD regions than the real data (recording session \#1: 55 cm vs. 229 cm, p=0.02, real vs. shuffled; recording session \#2: 72 cm vs. 231 cm, p=0.02, real vs. shuffled, one-sided; \FIGSUPP[4]{fig4sup1}). We also compared the HPD regions between the real data and the shuffle that preserved more of the local spatial structure while destroying global spatial structure. This also resulted in larger HPD regions than the real data for spatially coherent and incoherent SWRs (recording session \#1, spatially coherent: 29 cm vs. 52 cm, p=0.02, real vs. shuffled; recording session \#1, spatially incoherent: 181 cm vs. 243 cm, p=0.02, real vs. shuffled; recording session \#2, spatially coherent: 26 cm vs. 53 cm, p=0.02, real vs. shuffled; recording session \#2, spatially incoherent: 210 cm vs. 263 cm, p=0.02, real vs. shuffled; \FIGSUPP[4]{fig4sup2}).

@@

\subsection*{Control analyses}

@@

We then asked whether the slow spatially coherent events (stationary-continuous mixtures) potentially reflect a slow theta sweep rather than a replay event. We noted that this seems very unlikely given that the stationary-continuous mixtures tended to represent locations far from the animal that would only be seen during rapid theta sequences (median 50 cm \FIG{5}C). In addition, the majority of our events occur when the animal is moving at speeds < 2 cm/s, as expected for SWRs (\FIGSUPP[5]{fig5sup3}B). Further, when we restrict the analyses to SWRs where the animal was moving at speeds < 1 cm/s, we find similar proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}C).

@@

\subsection*{Comparison to standard approaches}

% Figure 6
\begin{figure}
\includegraphics[width=0.80\linewidth]{figures/Figure6/Figure6_v2}
\caption{
Comparison of the state space decoder to the standard "Bayesian" decoder. For each panel, the top five rows are the probability density of the estimated speed for SWRs that contain that dynamic or combination of dynamics. Note that these categories are not mutually exclusive since a SWR can have more than one dynamic or combination of dynamics. The sixth row is all SWRs containing unclassified dynamics. The final row is the estimated speed for all SWRs. \textbf{(A)} Radon transform  \textbf{(B)} MAP estimate  \textbf{(C)} Linear regression  \textbf{(D)} State space decoder presented in this manuscript.
}
\label{fig:6}
\end{figure}

Finally, we asked how our method compares to the standard approaches to identifying significant events and characterizing their content. This comparison is complicated due to the variety of parameters and inference methods used previously, so here we chose three commonly used approaches and a reasonable parameter set for the comparison. We began with an examination of events based on a measure of statistical significance, which has been the standard in the field.  We applied the clusterless decoder in 20 ms bins to produce probability densities of represented locations, as this binning is comparable to previous methods. We then used the radon transform method \citep{DavidsonHippocampalReplayExtended2009, FarooqEmergencepreconfiguredplastic2019} to calculate, for each SWR, a measure of significance based on the assumptions that the probability density moves at a constant speed throughout the event. We found that ~12\% of the classified events (using the full state-space model) were identified as significant. Not surprisingly, when events were classified based on which dynamic they contained, we found substantial variability: of 7176 SWRs that contained a period of stationary dynamics, 542 (8\%) were significant while of 4511 SWRs that contained a period of continuous dynamics, 1386 (~30\%) were significant. And when we restricted the analysis to stationary events that were >100 ms long, we found that 28\% were significant, and increase that is not surprising given that the shuffle used for the radon method penalizes shorter events. We also note that the linear regression method \citep{KarlssonAwakereplayremote2009, CarrHippocampalreplayawake2011, ShinDynamicsAwakeHippocampalPrefrontal2019}, detects only non-zero slopes as significant, and thus cannot identify stationary events as significant. The substantial variability in the proportions of "significant" events indicates that conclusions about which events are meaningful to the brain could be very different depending on the specific method and parameters used.

@@

We hypothesize that this large set of events supports the ability to store and update  representations of individual places and snippets of experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{YuDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to store or update the value of specific locations or short movement trajectories without having to generate full trajectories representing one of multiple ways to arrive at a location \citep{YuHippocampalcorticalinteraction2015}. Events that contain multiple dynamics (e.g. two continuous representations separated by a brief fragmented representation) might also help bind together experiences that occur in the same environment but at different places. Alternatively, it might be that downstream structures are attuned to these dynamics, and that each period of coherent hippocamapal dynamics drives its own set of responses elsewhere in the brain. We further note that recent results from a study of activity patterns during movement indicates that individual theta cycles can have different dynamics as well \citep{WangAlternatingsequencesfuture2020a}, suggesting that the underlying sequences seen during replay and theta may overlap in their content and dynamics.
